RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-21 manuscripts/manuscript.Rmd
Some edits from morning's chat
- Clarify scripted use vs interactive use in `get_trees`, `get_trees_list`
- Add future directions
- a few minor tweaks

Still need to revise introduction to emphasize semantics.

This paper has been written as a dynamic document 
[@Xie_2013].  The reader may paste the code blocks
shown into their R terminal to follow along with the exmples shown.

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
<!-- having a function's return type being more or less unpredictable -->
<!-- sounds like a bad design pattern. Is there at least precedence for -->
<!-- this practice in other R packages? HL -->

<!-- Yes, sadly it is.  Really this is a class definition issue that multiPhylo isn't a
proper generalization of a phylo object, but rather a different class entirely. 
Users who don't like this function can use the get_trees_list() to always be predictable.  -->

While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/) 
 through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```
<!-- the example makes me wonder how nexml_read() distinguishes -->
<!-- between a URL and a file name as its argument. HL -->
<!-- just type `nexml_read` and you'll find out ;-).  We use a simple regex  CB -->

@@

>>>>

This paper has been written as a dynamic document 
[@Xie_2013].  The reader may paste the code blocks
shown into their R terminal to follow along with the examples shown.

@@

Note that NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function is designed for interactive use, and thus attempts to return the most sensible R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.

For scripted use, it is desirable to have a consistent return type.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/) 
 through the
`nexml` format.  `RNeXML`'s `nexml_read` can also read a file directly from a URL by automatically recognizing the URL patterns and first downloading the file to a working directory:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```

@@

## Future Directions

For convenience, the RNeXML package can parse compressed XML binaries (gzipped XML) directly, facilitating work with very large data files.  As most R functions, especially those in common phylogenetics packages, require in-memory data structures, RNeXML currently loads this data directly into memory. Future work could facilitate manipulation of NeXML data directly from the disk.

Emerging formats such as the JSON based NeXSON format provide alternate representations of semantic, machine-readable phylogenetic data. At this time, R users must still rely on other tools that can convert these formats into NeXML.
Because the NeXML syntax is extensible, semantic, and machine-readable, this can be done without loss, unlike NEXUS formats.  Nevertheless, a possible future direction would be to provide direct support of such formats through RNeXML or a companion R package.
